Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-21 ind_files/06_unification_discussion.tex
Simplify and correct derivation of variance Price eqn
Move explanation of variance price from appendix C to discussion (chap 6)

&+ \underbrace{\frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}f_i\sqrt{A_i^+}dB_{t}^{(i)} - \overline{f}\sum\limits_{i=1}^{m}\sqrt{A_i^+}dB_{t}^{(i)}\right)}_{\text{Stochastic fluctuations}}

@@

&= \textrm{Var}(b^{\textrm{(ind)}}(\mathbf{x})) - \textrm{Var}(d^{\textrm{(ind)}}(\mathbf{x}))\label{nD_cross_covariance}

@@

where we have used that the covariance of a quantity with itself equals the variance (again, it is important to remember that all these variances and covariances are statistical properties of the population, and are not to be interpreted in the probabilistic sense). Thus, upon substituting $f = w$ in \eqref{nD_stochastic_Price} and taking expectations over the underlying probability space, we obtain:

@@

\negmedspace {\underbrace{\mystrut{4ex} \ \vphantom{ \frac{\textrm{Var}(b^{\textrm{(ind)}}) - \textrm{Var}(d^{\textrm{(ind)}})}{KN_K(t)} } \mathbb{E}\left[\textrm{Var}(w)\right] \ }_{\substack{\text{Fisher's} \\ \text{Fundamental theorem}}}} \ - \ {\underbrace{\mystrut{4ex}\mathbb{E}\left[\frac{\textrm{Var}(b^{\textrm{(ind)}}) - \textrm{Var}(d^{\textrm{(ind)}})}{KN_K(t)}\right]}_{\substack{\text{Noise-induced} \\ \text{selection}}}} \ + {\underbrace{\mystrut{4ex} \vphantom{ \frac{\textrm{Var}(b^{\textrm{(ind)}}) - \textrm{Var}(d^{\textrm{(ind)}})}{KN_K(t)} } \mathbb{E}\left[\hphantom{a}\overline{\hphantom{a}\frac{\partial w}{\partial t}\hphantom{a}}\hphantom{a}\right]}_{\substack{\text{Short-term (ecological)} \\ \text{feedbacks to fitness}}}}

@@

Taking $K \to \infty$ in \eqref{nD_stochastic_Fisher} recovers a well-known equation in population genetics upon noting that the process tends to a deterministic process as $K \to \infty$, as noted in section \ref{sec_nD_det_limit}, and thus the expectation value in the infinite population case is superfluous. The first term, $\textrm{Var}(w)$, is the subject of Fisher's fundamental theorem \citep{price_fishers_1972, frank_fishers_1992, kokko_stagnation_2021}. The last term arises in both finite and infinite populations whenever $w_i$ can vary over time \citep{baez_fundamental_2021}, be it through frequency-dependent selection, phenotypic plasticity, varying environments, or other ecological mechanisms, and represents feedback effects on the fitness $w_i$ of the $i\textsuperscript{th}$ species over short (`ecological') time-scales. The fact that Fisher appears to have been rather dismissive of this feedback, often only referring to it in a cursory manner as `environmental change' or `environmental degradation', has led to much discussion, debate, and confusion about the interpretation, importance, and implications of his `fundamental theorem' (see \cite{kokko_stagnation_2021} and sources cited therein). Finally, the second term of equation \eqref{nD_stochastic_Fisher} is a manifestation of noise-induced selection and vanishes in the infinite population limit, and is thus particular to finite populations.\\

@@

\mathbb{E}\left[\frac{d\overline{\tau}}{dt}\right] = \underbrace{\vphantom{ \mathbb{E}\left[\frac{\textrm{Var}(\tau)}{KN_K(t)}\right] } \mathbb{E}\left[\textrm{Var}(b^{\textrm{(ind)}}) - \textrm{Var}(d^{\textrm{(ind)}}\right]}_{\substack{\text{Classical selection} \\ \text{effects}}} \ - \ \underbrace{ \mathbb{E}\left[\frac{\textrm{Var}(\tau)}{KN_K(t)}\right]}_{\substack{\text{Noise-induced selection} \\ \text{effects}}} \ + \ \underbrace{ \vphantom{ \mathbb{E}\left[\frac{\textrm{Var}(\tau)}{KN_K(t)}\right] } \mathbb{E}\left[\hphantom{a}\overline{\hphantom{a}\frac{\partial \tau}{\partial t}\hphantom{a}}\hphantom{a}\right]}_{\substack{\text{Short-term (ecological)} \\ \text{feedbacks to $\tau_i$}}}

@@

The implications of this theorem have been extensively discussed in \cite{kuosmanen_turnover_2022}, which is where we refer the interested reader.

>>>>

&+ \underbrace{\frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f_i-\overline{f}\right)\sqrt{A_i^+}dB_{t}^{(i)}\right)}_{\text{Stochastic fluctuations}}

@@

&= \sigma^2_{b^{\textrm{(ind)}}(\mathbf{x})} - \sigma^2_{d^{\textrm{(ind)}}(\mathbf{x})}\label{nD_cross_covariance}

@@

where we have defined the statistical variance of a quantity $f$ as
\sigma^2_{f} \coloneqq \overline{(f^2)} - (\overline{f})^2
Note that just like the statistical mean, the statistical variance is a random variable and is not to be confused with the probabilistic/ensemble variance. Upon substituting $f = w$ in \eqref{nD_stochastic_Price} and taking expectations over the underlying probability space, we obtain:

@@

\negmedspace {\underbrace{\mystrut{4ex} \ \vphantom{ \frac{\sigma^2_{b^{\textrm{(ind)}}} - \sigma^2_{d^{\textrm{(ind)}}}}{KN_K(t)} } \mathbb{E}\left[\sigma^2_{w}\right] \ }_{\substack{\text{Fisher's} \\ \text{Fundamental theorem}}}} \ - \ {\underbrace{\mystrut{4ex}\mathbb{E}\left[\frac{\sigma^2_{b^{\textrm{(ind)}}} - \sigma^2_{d^{\textrm{(ind)}}}}{KN_K(t)}\right]}_{\substack{\text{Noise-induced} \\ \text{selection}}}} \ + {\underbrace{\mystrut{4ex} \vphantom{ \frac{\sigma^2_{b^{\textrm{(ind)}}} - \sigma^2_{d^{\textrm{(ind)}}}}{KN_K(t)} } \mathbb{E}\left[\hphantom{a}\overline{\hphantom{a}\frac{\partial w}{\partial t}\hphantom{a}}\hphantom{a}\right]}_{\substack{\text{Short-term (ecological)} \\ \text{feedbacks to fitness}}}}

@@

Taking $K \to \infty$ in \eqref{nD_stochastic_Fisher} recovers a well-known equation in population genetics upon noting that the process tends to a deterministic process as $K \to \infty$, as noted in section \ref{sec_nD_det_limit}, and thus the expectation value in the infinite population case is superfluous. The first term, $\sigma^2_w$, is the subject of Fisher's fundamental theorem \citep{price_fishers_1972, frank_fishers_1992, kokko_stagnation_2021}. The last term arises in both finite and infinite populations whenever $w_i$ can vary over time \citep{baez_fundamental_2021}, be it through frequency-dependent selection, phenotypic plasticity, varying environments, or other ecological mechanisms, and represents feedback effects on the fitness $w_i$ of the $i\textsuperscript{th}$ species over short (`ecological') time-scales. The fact that Fisher appears to have been rather dismissive of this feedback, often only referring to it in a cursory manner as `environmental change' (or `environmental degradation'), has led to much discussion, debate, and confusion about the interpretation, importance, and implications of his `fundamental theorem' (see \cite{kokko_stagnation_2021} and sources cited therein). Finally, the second term of equation \eqref{nD_stochastic_Fisher} is a manifestation of noise-induced selection and vanishes in the infinite population limit, and is thus particular to finite populations.\\

@@

\mathbb{E}\left[\frac{d\overline{\tau}}{dt}\right] = \underbrace{\vphantom{ \mathbb{E}\left[\frac{\sigma^2_{\tau}}{KN_K(t)}\right] } \mathbb{E}\left[\sigma^2_{b^{\textrm{(ind)}}} - \sigma^2_{d^{\textrm{(ind)}}}\right]}_{\substack{\text{Classical selection} \\ \text{effects}}} \ - \ \underbrace{ \mathbb{E}\left[\frac{\sigma^2_{\tau}}{KN_K(t)}\right]}_{\substack{\text{Noise-induced selection} \\ \text{effects}}} \ + \ \underbrace{ \vphantom{ \mathbb{E}\left[\frac{\sigma^2_{\tau}}{KN_K(t)}\right] } \mathbb{E}\left[\hphantom{a}\overline{\hphantom{a}\frac{\partial \tau}{\partial t}\hphantom{a}}\hphantom{a}\right]}_{\substack{\text{Short-term (ecological)} \\ \text{feedbacks to $\tau_i$}}}

@@

The implications of this theorem have been extensively discussed in \citep{kuosmanen_turnover_2022}, which is where we refer the interested reader.
\hl{Not sure if there's much more to write?}

\section{Fundamental theorems for the variance of quantities}
Equation \eqref{nD_stochastic_Price} is a general equation for the mean value of an arbitrary type level quantity $f$ in the population. In many real-life situations, especially those pertaining to finite populations, we are interested in not just the mean, but also the variance of a type-level quantity. In appendix \ref{App_stoch_var_eqns}, we show that the statistical variance of any type level quantity $f$ obeys
\label{stochastic_Price_variance}
d\sigma^2_{f} &= \textrm{Cov}\left(w,(f - \overline{f})^2\right)dt - \frac{1}{KN_K}\left[ \ \overline{\tau}\sigma^2_{f} +  2\textrm{Cov}\left(\tau,(f - \overline{f})^2\right) \ \right]dt\\[12pt]
& + 2\textrm{Cov}\left(\frac{\partial f}{\partial t},f\right)dt + M_{\sigma^2_f}(\mathbf{p},N_K)dt + dB_{\sigma^2_{f}}
where
\label{variance_price_mutation_term}
M_{\sigma^2_f}(\mathbf{p},N_K) = \mu\left(\sum\limits_{i=1}^{m}\left[\left(1 - \frac{2}{KN_K}\right)(f_i - \overline{f})^2Q_i(\mathbf{p})\right] + \sigma^2_f\left(1 - \frac{1}{KN_K}\right)\sum\limits_{i=1}^{m}Q_i(\mathbf{p})\right)
is a mutational term that vanishes in the $\mu \to 0$ limit and
\label{variance_price_diffusion_term}
dB_{\sigma^2_f} = \frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2\sqrt{A_i^+}dB_{t}^{(i)}\right)
is a stochastic integral term measuring the (non-directional) effect of stochastic fluctuations that vanishes upon taking an expectation over the probability space. In the case of one-dimensional quantitative traits, an infinite-dimensional version of \eqref{stochastic_Price_variance} has recently been rigorously derived \citep{week_white_2021} using measure-theoretic tools under certain additional assumptions - Equation (21b) of \cite{week_white_2021} is precisely the $m \to \infty$ version of my equation \eqref{stochastic_Price_variance} when $f$ is the phenotypic value of the quantitative trait under consideration (and thus $\partial f/\partial t \equiv 0$), the mutation is modeled through a Gaussian mutational kernel, and the $dB_{\sigma^2_f}$ term is suitably reformulated as a stochastic integral against a single space-time white noise process.\\
Once again, terms of equation \eqref{stochastic_Price_variance} lend themselves to straightforward biological interpretation. The quantity $(f-\overline{f})^2$ is a measure of the `spread' of the quantity $f$ around the population mean value $\overline{f}$, and thus covariance with $(f-\overline{f})^2$ is a measure of disruptive (\emph{i.e.} diversifying) selection: If values of $f$ that are further away from the mean value are associated with higher fitness (resp. lower $\tau$), the variance of that quantity in the population increases. An extreme case is evolutionary branching, where the mean value $\overline{f}$ is at a local minima for fitness but $f_i \not\equiv \overline{f}$ and thus $\textrm{Cov}(w,(f - \overline{f})^2)$ is strongly positive. In this case, the population experiences a sudden explosion in variance, and, if $\textrm{Cov}(w,(f - \overline{f})^2)$ is still positive after the initial emergence of multiple morphs, evolution eventually leads to the emergence of stable coexisting polymorphisms in the population. The $\textrm{Cov}\left(\partial f/\partial t,f\right)$ term once again represents the effect of eco-evolutionary feedback loops due to rapid change in $f$ that is not solely due to changes in $\mathbf{p}$. The $M_{\sigma^2_f}(\mathbf{p},N_K)$ term quantifies the effect of mutations on the variance of $f$. Note that each $Q_i(\mathbf{p}) \geq 0$ by its definition in \eqref{nD_functional_forms_for_replicator}, and the total population size $KN_K \geq 2$ for all interesting evolutionary questions. Thus, from \eqref{variance_price_mutation_term}, it is clear that $M_{\sigma^2_f}(\mathbf{p},N_K) \geq 0$, meaning that mutations never decrease the variance of $f$ in the population.\\
The $\overline{\tau}\sigma^2_{f}$ term represents a loss of diversity due to stochastic extinctions (i.e. demographic stochasticity).To see this, let us consider the case in which this is the only force at play. Let us imagine we have a population of asexual organisms and each $f_i$ is simply a label or mark arbitrarily assigned to individuals in the population at the start of an experiment/observational study and subsequently passed on to offspring - For example, a neutral genetic tag in a part of the genome that experiences a negligible mutation rate. Let us set $\mu = 0$ so that the labels cannot change between parents to offspring. This means that we have $M_{\sigma^2_f}(\mathbf{p},N_K) \equiv 0$. Further, since the labels are arbitrary and have no effect whatsoever on the biology of the organisms, we have $\textrm{Cov}\left(w,(f - \overline{f})^2\right) \equiv \textrm{Cov}\left(\tau,(f - \overline{f})^2\right) \equiv 0$. Since the labels do not change over time, we also have $\textrm{Cov}\left(\partial f/\partial t,f\right) = 0$. From \eqref{stochastic_Price_variance}, we see that in this case, the variance changes as
d\sigma^2_f = - \frac{\overline{\tau}\sigma^2_{f}}{KN_K}dt + dB_{\sigma^2_{f}}
On taking expectations, we see that the expected variance in the population obeys
\label{neutral_example_for_variances}
\frac{d \mathbb{E}[\sigma^2_f]}{dt} = - \left(\mathbb{E}\left[\frac{\overline{\tau}}{KN_K}\right]\right)\mathbb{E}[\sigma^2_{f}]
We were able to decompose the expectation on the RHS because the label $f$ is completely arbitrary and thus independent of both $\overline{\tau}$ and $N_K(t)$. Equation \eqref{neutral_example_for_variances} is a simple linear ODE and has the solution
\mathbb{E}[\sigma^2_f(t)] = \sigma^2_f(0)\exp\left(-\mathbb{E}\left[\frac{\overline{\tau}}{KN_K}\right]\right)
which tells us that the population is expected to lose its diversity of labels at an exponential rate. Populations with higher mean turnover $\overline{\tau}$  and/or lower population size $KN_K$ lose diversity faster, which should be intuitive since these populations experience more extreme stochastic events and are therefore more likely to experience stochastic loss of labels. Note that \emph{which} labels/individuals are lost is entirely random (since all labels are arbitrary), but nevertheless, labels tend to be stochastically lost until only a small number of labels remain in the population.

### 2023-02-21 ind_files/99_Appendix_C.tex
Move explanation of variance price from appendix C to discussion (chap 6)

We will evaluate the RHS term by term. The first term is as simplified as can be. For the second term, we can substitute $dp_i$ from \eqref{nD_stochastic_RM} and use the same steps used in going from \eqref{nD_replicator_mutator} to \eqref{nD_Price} to write

@@

&+\frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}f^2_i\sqrt{A_i^+}dB_{t}^{(i)} - \overline{f^2}\sum\limits_{i=1}^{m}\sqrt{A_i^+}dB_{t}^{(i)}\right)

@@

where the relevant functions $S$ and $D_j$ can be read off from \eqref{nD_stochastic_Price}. Since the $dB$ terms are unwieldy and do not contribute to the expected dynamics, we will denote the contribution of all the $dB_t$ terms collectively by $dB_{\sigma^2_{f}}$ and not explicitly calculate these terms. We also note that the covariance operator is a bilinear form, \emph{i.e.} given any three quantities $X$, $Y$ and $Z$ and any constant $a \neq 0$, we have the relations:

@@

&= \frac{1}{KN_K}\left(\overline{\left(f - \overline{f}\right)^2\tau}+\mu \sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2Q_i(\mathbf{p})\right)\\

@@

&= \frac{1}{KN_K}\left(\textrm{Cov}(\tau,\left(f - \overline{f}\right)^2) + \overline{\tau}\sigma^2_{f} +\mu \sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2Q_i(\mathbf{p})\right)

@@

where we have used the definition of statistical covariance in the second to last line and used the definition of statistical variance in the last line. Substituting this into \eqref{intermediate_1_for_variances} and using $M_{\sigma^2_f}(\mathbf{p},N_K)$ to denote the contributions of all the mutational terms (\emph{i.e.} all terms with a $\mu$ factor) for notational brevity, we obtain

@@

d\sigma^2_{f} &= \textrm{Cov}(w,f^2 - 2\overline{f}f)dt - \frac{1}{KN_K}\left(\textrm{Cov}(\tau,f^2 - 2\overline{f}f) + \textrm{Cov}(\tau,\left(f - \overline{f}\right)^2) + \overline{\tau}\sigma^2_{f}\right)dt\\
&+ 2\textrm{Cov}\left(f, \frac{\partial f}{\partial t}\right)dt + M_{\sigma^2_f}(\mathbf{p},N_K)dt + dB_{\sigma^2_{f}}

@@

\label{intermediate_2_for_variances}

@@

and thus, using this in \eqref{intermediate_2_for_variances},  we see that the rate of change of the variance of any type-level quantity $f$ in the population satisfies:

@@

\label{stochastic_Price_variance}

@@

In the case of one-dimensional quantitative traits, an infinite-dimensional version of \eqref{stochastic_Price_variance} has been rigorously derived \citep{week_white_2021} in a recent paper for the case where $f$ is the phenotypic value of the quantitative trait - Equation (21b) of \cite{week_white_2021} is precisely the $m \to \infty$ version of my equation \eqref{stochastic_Price_variance} when $f$ is the phenotypic value of the quantitative trait under consideration.\\
Once again, terms of equation \eqref{stochastic_Price_variance} lend themselves to straightforward biological interpretation. The quantity $(f-\overline{f})^2$ is a measure of the `spread' of the quantity $f$ around the population mean value $\overline{f}$, and thus covariance with $(f-\overline{f})^2$ is a measure of disruptive (\emph{i.e.} diversifying) selection: If values of $f$ that are further away from the mean value are associated with higher fitness, the variance of that quantity in the population increases. The $\overline{\tau}\sigma^2_{f}$ term represents a loss of diversity due to stochastic extinctions (i.e. demographic stochasticity). To see this, let us consider the case in which each $f_i$ is simply a label or mark arbitrarily assigned to individuals in the population at the start of an experiment/observation and subsequently passed on to offspring - For example, a neutral tag in the genome of the organism in an area that does not mutate. Let us set $\mu = 0$ so that the labels cannot change between parents to offspring. This means that we have $M_{\sigma^2_f}(\mathbf{p},N_K) \equiv 0$. Further, since the labels are arbitrary and have no effect whatsoever on the biology of the organisms, we have $\textrm{Cov}\left(w,(f - \overline{f})^2\right) \equiv \textrm{Cov}\left(\tau,(f - \overline{f})^2\right) \equiv 0$. Since the labels do not change over time, we also have $\textrm{Cov}\left(\frac{\partial f}{\partial t},f\right) = 0$. From \eqref{stochastic_Price_variance}, we see that in this case, the variance changes as

@@

d\sigma^2_f = - \frac{\overline{\tau}\sigma^2_{f}}{KN_K}dt + dB_{\sigma^2_{f}}

@@

In particular, on taking expectations, we see that the expected variance in the population obeys

@@

\frac{d \mathbb{E}[\sigma^2_f]}{dt} = - \left(\mathbb{E}\left[\frac{\overline{\tau}}{KN_K}\right]\right)\mathbb{E}[\sigma^2_{f}]

@@

We were able to decompose the expectation on the RHS because the label $f$ is completely arbitrary and thus independent of both $\overline{\tau}$ and $N_K(t)$. This is a simple linear ODE and has the solution

@@

\mathbb{E}[\sigma^2_f(t)] = \sigma^2_f(0)\exp\left(-\mathbb{E}\left[\frac{\overline{\tau}}{KN_K}\right]\right)

@@

In this case, demographic stochasticity is expected to lead to an exponential loss of diversity in the population. Populations with higher mean turnover $\overline{\tau}$  and/or lower population size $KN_K$ lose diversity faster, which again should be intuitive since these populations experience more extreme stochastic events and are therefore more likely to experience loss of labels. Note that \emph{which} label is lost is entirely random (since labelling is arbitrary), but nevertheless, labels tend to be stochastically lost until only a small number of labels remain in the population.

>>>>

We will evaluate the RHS term by term. The first term is as simplified as can be without more information about $f$. For the second term, we can substitute $dp_i$ from \eqref{nD_stochastic_RM} and use the same steps used in going from \eqref{nD_replicator_mutator} to \eqref{nD_Price} to write

@@

&+\frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f^2_i - \overline{f^2}\right)\sqrt{A_i^+}dB_{t}^{(i)}\right)

@@

where the relevant functions $S$ and $D_j$ can be read off from \eqref{nD_stochastic_Price}. Since the $dB$ terms are unwieldy, we will denote the contribution of all the $dB_t$ terms collectively by $dB_{\sigma^2_{f}}$ to reduce notational clutter and only explicitly calculate these terms at the end. We also note that the covariance operator is a bilinear form, \emph{i.e.} given any three quantities $X$, $Y$ and $Z$ and any constant $a \neq 0$, we have the relations:

@@

&= \frac{1}{KN_K}\left(\overline{\tau\left(f - \overline{f}\right)^2}+\mu \sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2Q_i(\mathbf{p})\right)\\

@@

&= \frac{1}{KN_K}\left(\textrm{Cov}(\tau,\left(f - \overline{f}\right)^2) + \overline{\tau}\sigma^2_{f} +\mu \sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2Q_i(\mathbf{p})\right)\label{intermediate_2_for_variances}

@@

where we have used the definition of statistical covariance in the second to last line and used the definition of statistical variance in the last line. Substituting \eqref{intermediate_2_for_variances} into \eqref{intermediate_1_for_variances} and using $M_{\sigma^2_f}(\mathbf{p},N_K)$ to denote the contributions of all the mutational terms (\emph{i.e.} all terms with a $\mu$ factor) for notational brevity, we obtain

@@

d\sigma^2_{f} &= \textrm{Cov}(w,f^2 - 2\overline{f}f)dt - \frac{1}{KN_K}\left(\textrm{Cov}(\tau,f^2 - 2\overline{f}f) + \textrm{Cov}(\tau,\left(f - \overline{f}\right)^2) + \overline{\tau}\sigma^2_{f} \right)dt\\
&+ 2\textrm{Cov}\left(\frac{\partial f}{\partial t},f\right)dt + M_{\sigma^2_f}(\mathbf{p},N_K)dt + dB_{\sigma^2_{f}}

@@

\label{intermediate_3_for_variances}

@@

and thus, using this in \eqref{intermediate_3_for_variances},  we see that the rate of change of the variance of any type-level quantity $f$ in the population satisfies:

@@

\label{app_stochastic_Price_variance}

@@

This is precisely equation \eqref{stochastic_Price_variance} in the main text. To calculate the mutation term, we substitute \eqref{intermediate_2_for_variances} into \eqref{intermediate_1_for_variances} to find

@@

M_{\sigma^2_f}(\mathbf{p},N_K) &= \mu\left(\sum\limits_{i=1}^{m}\left(f^2_i - 2\overline{f}f_i -\overline{f^2}+2\overline{f}^2\right)Q_i(\mathbf{p})\right)\\
&- \frac{\mu}{KN_K}\sum\limits_{i=1}^{m}\left(f^2_i - 2\overline{f}f_i -\overline{f^2}+2\overline{f}^2 + (f_i - \overline{f})^2\right)Q_i(\mathbf{p}) 

@@

We can further simplify the first term of the RHS as
f^2_i - 2\overline{f}f_i -\overline{f^2}+2\overline{f}^2 &= (f^2_i +\overline{f}^2 - 2\overline{f}f_i) - (\overline{f^2} - \overline{f}^2)\\
&= (f_i - \overline{f})^2 + \sigma^2_f
and similarly, the second term as
f^2_i - 2\overline{f}f_i -\overline{f^2}+2\overline{f}^2 + (f_i - \overline{f})^2 = 2(f_i - \overline{f})^2 + \sigma^2_f
thus, the contributions of mutations to the change in the variance of $f$ are given by

@@

M_{\sigma^2_f}(\mathbf{p},N_K) &= \mu\left(\sum\limits_{i=1}^{m}\left((f_i - \overline{f})^2 + \sigma^2_f\right)Q_i(\mathbf{p})\right)\\
&- \frac{\mu}{KN_K}\sum\limits_{i=1}^{m}\left(2(f_i - \overline{f})^2 + \sigma^2_f\right)Q_i(\mathbf{p}) 

@@

which after slight rearrangement becomes

@@

M_{\sigma^2_f}(\mathbf{p},N_K) = \mu\left(\sum\limits_{i=1}^{m}\left[\left(1 - \frac{2}{KN_K}\right)(f_i - \overline{f})^2Q_i(\mathbf{p})\right] + \sigma^2_f\left(1 - \frac{1}{KN_K}\right)\sum\limits_{i=1}^{m}Q_i(\mathbf{p})\right)

@@

which is equation \eqref{variance_price_mutation_term} in the main text. For the $dB$ terms, we can use equations \eqref{2nd_term_for_variances} and \eqref{3rd_term_for_variances} to calculate:
dB_{\sigma^2_{f}} &=  \frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f^2_i - \overline{f^2} - 2 \overline{f}(f_i - \overline{f})\right)\sqrt{A_i^+}dB_{t}^{(i)}\right)\\
&= \frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f^2_i - \overline{f^2} - 2 \overline{f}f_i - 2\overline{f}^2\right)\sqrt{A_i^+}dB_{t}^{(i)}\right)\\
&= \frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2\sqrt{A_i^+}dB_{t}^{(i)}\right)
which is equation \eqref{variance_price_diffusion_term} in the main text.
